Unifying intra- and inter-specific variation in tropical tree mortality

### 2017-12-04 ms/manuscript.Rnw
final set up fixes

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalized the article contents.
\verticaladjustment{-2pt}

@@

\subsection*{Hazard functions}
Comparing the three hazard functions in Fig. \ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only or growth-dependent only functions (Fig. \ref{fig2}A). Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig. \ref{fig2}A). A summary of hyper parameter estimates can be found in Table S1.

@@

\subsection*{Parameter contribution to predicted variation}
To compliment the above analysis based on skill in predicting held-out (test) data, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1yr for all individuals (Fig. \ref{fig5}). We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model. Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality. Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation. Census accounted for only \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation. The growth-dependent hazard (i.e. the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig. \ref{fig5}).

@@

Eqs. \ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census and wood density, as well as random effects. Growth rates were estimated from field measurements of diameter, which inevitably include observation error. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_dbh)}\% of estimated growth rates were negative. To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig. S1). The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of both wood density (measured at species level) and a species-level random effect:
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$, modelled as random realisations from log-normal distributions. The form of eq. \ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around $\rho_c$. We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low. Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details). Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see Supplementary Material S3-S4 for details). We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior. Species parameter estimates from the final model are shown in Figs. S3\textendash S5.

@@

\subsection*{Posthoc correlations}
We calculated a gap index as a measure of a species' light dependence using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995. The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m). For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap). As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots. Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1. This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots. These weighted values were then summed for each central cell and then these values were normalised to be between 0 and 1. The gap index for each species was then estimated as the average light level encountered by new saplings appearing in the census (Fig. S2).

>>>>

\dates{This manuscript was compiled on \today}
%\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\dates{This manuscript was compiled on \today}
%\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalized the article contents.
\verticaladjustment{-2pt}

@@

\subsection*{Hazard functions}
Comparing the three hazard functions in Fig. \ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only (i.e. null) or growth-dependent only functions (Fig. \ref{fig2}A). Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig. \ref{fig2}A). A summary of hyper parameter estimates can be found in Table S1.

@@

\subsection*{Parameter contribution to predicted variation}
To compliment the analysis, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1 year for all individuals (Fig. \ref{fig5}). We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model. Species effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality. Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation. Census accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation. The growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig. \ref{fig5}).

@@

Eqs. \ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census and wood density, as well as random effects. Growth rates were estimated from field measurements of diameter, which inevitably include observation error. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_dbh)}\% of estimated growth rates were negative. To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig. S1). The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of both wood density (measured at species level) and a species-level random effect:
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$. These random effects were modelled as random realisations from log-normal distributions. The form of eq. \ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around $\rho_c$. We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low. Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details). Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see Supplementary Material S3-S4 for details). We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior. Species parameter estimates from the final model are shown in Figs. S3\textendash S5.

@@

\subsection*{Posthoc correlations}
We calculated a gap index as a measure of a species' light dependence using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995. The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m). For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap). As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots. Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1. This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots. These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots. The gap index for each species was estimated as the mean light index encountered by new saplings appearing in the census (Fig. S2).
